\MakeUppercase{The Structure of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-07-04 k-planar.tex
Revised page 1

\begin{lem}\lemlabel{nice-decomposition}
  Let $H$ be a graph of treewidth $t-1$.  Then there exists a graph $H'$ isomorphic to $H$ that has a tree-decomposition $(B_x:x\in V(T))$ of width at most $t-1$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    \item\tlabel{node-set} $V(H')\subseteq V(T)$.
    \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each $v'\in V(H')$, the subtree $T[v']:=T[\{x\in V(T):v'\in B_x\}]$ is rooted at $v'$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $v'w'\in E(H)$, one of $v'$ or $w'$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

\begin{proof}
  Begin with a width-$(t-1)$ tree-decomposition $(C_x:x\in V(T_0))$ of $H$ using a tree $T_0$ with the property that, for any edge $xy\in E(T_0)$, $|C_x\ominus C_y|\le 1$. That is, $C_x$ contains exactly one vertex not in $C_y$ or vice-versa. It is well-known that such a tree-decomposition exists.

  Next, select an arbitrary node $x\in V(T_0)$ and let $C_x:=\{v_1,\ldots,v_{t'}\}$ denote the contents of $C_x$.
  Create a new tree $T$ from $T_0$ by adding a path of nodes $x_0,x_1,x_2,\ldots,x_{t'-1},x_{t'}=x$. Let $C_{x_0}:=\emptyset$ and, for each $i\in\{1,\ldots,t'\}$, $C_{x_i}:=\{v_1,\ldots,v_i\}$.  Note that $(C_x: x\in V(T))$ is a width-$(t-1)$ tree-decomposition of $H$. Let $T$ be rooted at $x_0$. Note that $(C_x:x\in V(T))$ satisfies Properties~\tref{rooted}, and \tref{diff}.  For the remaining properties some renaming is required, which we now describe.

  For each $v\in V(H)$, let $v'\in V(T)$ be the unique node such that $v\in C_{v'}$ and $v'$ has minimum $T$-depth.  The uniqueness of $v'$ follows from the fact that $T[v]:=T[\{x\in V(T):v\in B_x\}]$ is connected, so $v'$ is the root of $T[v]$.  Furthermore, for any pair of distinct vertices $v,w\in V(H)$, \tref{rooted} and \tref{diff} ensure that $v'\neq w'$ since, otherwise, $C_{v'}\ominus C_{x'}\supseteq \{v',w'\}$ where $x'$ is the parent of $v'=w'$.  Property~\tref{rooted} would ensure that $x'$ exists and Property~\tref{diff} would ensure that $|C_{x'}\ominus C_{v'}|\le 1$.

  For each $x\in V(T)$, let $B_x=\{v' : v\in C_x\}$.  Now define the graph $H'$ to have vertex set $V(H')=\{v':v\in V(H)\}$ and edge set $E(H')=\{v'w': vw\in E(H)\}$.  Obviously, $H'$ isomorphic to $H$ and $V(H')\subseteq V(T)$ and $(B_x: x\in V(T))$ is a width=$(t-1)$ tree-decomposition of $H'$.  The decomposition $(B_x: x\in V(T))$ inherits Properties~\tref{rooted} and \tref{diff} from $(C_x:x\in V(T))$ and satisfies Property~\tref{node-set} and \tref{subtree-root} by definition.

  To see that $(B_x:x\in V(T))$ satisifies Property~\tref{ancestor-edge}, observe that, if $vw\in E(H)$, then at least one of $v'$ or $w'$ is contained in $B_x$ for every node $x$ on the path from $v'$ to $w'$ in $T$.  If neither $v'$ nor $w'$ is an ancestor of the other, then some node $x$ on this path has $T$-depth less than that of $v'$ and $w'$.  If $v'\in B_x$ this contradicts the fact that $v'$ is the root of $T[v']$.  If $w'\in B_x$ this contradicts the fact that $w'$ is the root of $T[w']$.
\end{proof}

@@

It is natural to ask if \thmref{modern-major-general} is tight.  We do not answer this question, but we do point out that, even when $t=1$ and $d=3$, an exponential dependence on $k$ is necessary.  If $G$ is a complete binary tree on $2^{k/2}$ leaves, then $G^{k,3}$ contains a clique $K$ of size  $s=2^{k/2+1}-1$.  If these vertices appear in $r$ parts of the $I$-partition, then the treewidth of $I$ is at least $r-1$.  On the other hand, these vertices must all appear on two consecutive layers, so the layered width of the partition is at least $s/(2r)$. Therefore any layered $I$ partition of $G^{k,d}$ of layered width $\ell'$ in which the treewidth of $I$ is at most $t'-1$ must satisfy
\[
    \ell'\times t' \ge 2^{k/2}-1/2 \enspace .
\]

\newpage

>>>>

\begin{lem}\lemlabel{nice-decomposition}
  Let $H$ be a graph of treewidth $t$.  Then there exists a graph $H'$ isomorphic to $H$ that has a tree-decomposition $(B_x:x\in V(T))$ of width at most $t$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    \item\tlabel{node-set} $V(H')\subseteq V(T)$.
    \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each $v'\in V(H')$, the subtree $T[v']:=T[\{x\in V(T):v'\in B_x\}]$ is rooted at $v'$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $v'w'\in E(H)$, one of $v'$ or $w'$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

\begin{proof}
  Begin with a width-$t$ tree-decomposition $(C_x:x\in V(T_0))$ of $H$ using a tree $T_0$ with the property that, for any edge $xy\in E(T_0)$, $|C_x\ominus C_y|\le 1$. That is, $C_x$ contains exactly one vertex not in $C_y$ or vice-versa. It is well-known that such a tree-decomposition exists.  Indeed, any tree-decomposition of $H$ can be made to have this property adding $|C_x\ominus C_y|-1$ subdivision vertices along each edge $xy\in E(T_0)$ and assigning approriate bags to the newly introduced subdivision vertices.

  Next, select an arbitrary node $x\in V(T_0)$ and let $\{v_1,\ldots,v_{t'}\}:=C_x$ denote the contents of $C_x$. Create a new tree $T$ from $T_0$ by adding a path of nodes $x_0,x_1,x_2,\ldots,x_{t'-1},x_{t'}=x$. Let $C_{x_0}:=\emptyset$ and, for each $i\in\{1,\ldots,t'\}$, $C_{x_i}:=\{v_1,\ldots,v_i\}$.  Root $T_0$ at $x_0$ and note that $(C_x: x\in V(T))$ is a width-$t$ tree-decomposition of $H$ that satisfies Properties~\tref{rooted}, and \tref{diff}.  For the remaining properties some renaming is required, which we now describe.

  For each $v\in V(H)$, let $v'\in V(T)$ be the unique node such that $v\in C_{v'}$ and $v'$ has minimum $T$-depth.  The uniqueness of $v'$ follows from the fact that $T[v]:=T[\{x\in V(T):v\in B_x\}]$ is connected, so $v'$ is the root of $T[v]$.  Furthermore, for any pair of distinct vertices $v,w\in V(H)$, \tref{rooted} and \tref{diff} ensure that $v'\neq w'$: Otherwise, $C_{v'}\ominus C_{x'}\supseteq \{v',w'\}$ where $x'$ is the parent of $v'=w'$.  Property~\tref{rooted} ensures that $x'$ exists but then Property~\tref{diff} contradicts $C_{v'}\ominus C_{x'}\supseteq \{v',w'\}$.

  For each $x\in V(T)$, let $B_x=\{v' : v\in C_x\}$.  Now define the graph $H'$ to have vertex set $V(H')=\{v':v\in V(H)\}$ and edge set $E(H')=\{v'w': vw\in E(H)\}$.  Obviously, $H'$ isomorphic to $H$, $V(H')\subseteq V(T)$ and $(B_x: x\in V(T))$ is a width-$t$ tree-decomposition of $H'$.  The decomposition $(B_x: x\in V(T))$ inherits Properties~\tref{rooted} and \tref{diff} from $(C_x:x\in V(T))$ and satisfies \tref{node-set} and \tref{subtree-root} by definition.

  To see that $(B_x:x\in V(T))$ satisifies \tref{ancestor-edge}, observe that, if $vw\in E(H)$, then at least one of $v'$ or $w'$ is contained in $B_x$ for every node $x$ on the path from $v'$ to $w'$ in $T$.  If neither $v'$ nor $w'$ is an ancestor of the other, then some node $x$ on this path has $T$-depth less than that of $v'$ and $w'$.  If $v'\in B_x$ this contradicts the fact that $v'$ is the root of $T[v']$.  If $w'\in B_x$ this contradicts the fact that $w'$ is the root of $T[w']$.
\end{proof}

@@

It is natural to ask if \thmref{modern-major-general} is tight.  We do not answer this question, but we do point out that, even when $t=1$ and $d=3$, an exponential dependence on $k$ is necessary.  If $G$ is a complete binary tree on $2^{k/2}$ leaves, then $G^{k,3}$ contains a clique $K$ of size  $s=2^{k/2+1}-1$.  If these vertices appear in $r$ parts of the $I$-partition, then the treewidth of $I$ is at least $r-1$.  On the other hand, these vertices must all appear on two consecutive layers, so the layered width of the partition is at least $s/(2r)$. Therefore any layered $I$ partition of $G^{k,d}$ of layered width $\ell'$ in which the treewidth of $I$ is at most $t'-1$ must satisfy
\[
    \ell'\times t' \ge 2^{k/2}-1/2 \enspace .
\]   
\newpage
